A latency-tolerant runtime for large graph computations

### 2011-01-17 doc/hotpar2011/softxmt-hotpar2011.tex
minor formatting edits

Multithreading is a technique that has been used successfully to
implement efficient computations for these graphs. The Cray XMT is an
example of such a solution: it tries to solve the memory latency
problem through concurrency rather than caching. Each XMT processor
supports 128 hardware contexts and $~$1000 outstanding memory
operations and is able to switch contexts every cycle. But this
approach has a limitation: each context is significantly slower than a
modern commodity processor, yielding bad single-threaded
performance. While the XMT is fast at sparse, low-degree graph
problems, it is slow at other scientific codes that can take advantage
of commodity processors' memory hierarchies. Thus, few XMTs have been
built.

@@

Though not strictly necessary, for ease of exposition we imagine a
one-to-one binding between threads and cores.  Our model manifests
within any thread independently of the others as a ``fray'' [[for lack
of better term ]].  Within the thread, a fray instance is the
instantiation of a set of coroutines and a scheduler.  Instantiation
is at user-level and the operating system sees the entire fray as as a
single user-level thread.

[[ describe calls that perform thread & scheduler instantiation; and schedule activation ]]

@@

[[ describe calls to prefetch-and-yield ]]

@@

[[ discussion of coroutines & synchronization ]] Were a coroutine to
block on a synchronization variable shared with other threads, the
entire fray would suspend execution.  This can lead to deadlock when,
for example, one coroutine waits to consume from another thread that
is waiting to consume what only another coroutine in this first fray
can produce.  Instead, coroutines must yield on failed synchronization
events, spinning rather than blocking, where were they bonafide
thereads, blocking might be more efficient.

[[ say something about ``synchronization'' within a fray ]]

>>>>

Multithreading is a technique that has been used successfully to
implement efficient computations for these graphs. The Cray XMT is an
example of such a solution: it tries to solve the memory latency
problem through concurrency rather than caching. Each XMT processor
supports 128 hardware contexts and $\sim$1000 outstanding memory
operations and is able to switch contexts every cycle. But this
approach has a limitation: each context is significantly slower than a
modern commodity processor, yielding bad single-threaded
performance. While the XMT is fast at sparse, low-degree graph
problems, it is slow at other scientific codes that can take advantage
of commodity processors' memory hierarchies. Thus, few XMTs have been
built.

@@

Though not strictly necessary, for ease of exposition we imagine a
one-to-one binding between threads and cores.  Our model manifests
within any thread independently of the others as a ``fray'' \todo{for lack
of better term }.  Within the thread, a fray instance is the
instantiation of a set of coroutines and a scheduler.  Instantiation
is at user-level and the operating system sees the entire fray as as a
single user-level thread.

\todo{ describe calls that perform thread \& scheduler instantiation; and schedule activation }

@@

\todo{ describe calls to prefetch-and-yield }

@@

\todo{ discussion of coroutines \& synchronization } Were a coroutine to
block on a synchronization variable shared with other threads, the
entire fray would suspend execution.  This can lead to deadlock when,
for example, one coroutine waits to consume from another thread that
is waiting to consume what only another coroutine in this first fray
can produce.  Instead, coroutines must yield on failed synchronization
events, spinning rather than blocking, where were they bonafide
thereads, blocking might be more efficient.

\todo{ say something about ``synchronization'' within a fray }
